---
title: "Practical Machine Learning"
author: "Courtney Rowlands"
date: "10/14/2019"
output: 
  html_document:
    keep_md: yes
pdf_document: default
---

<!-- rmarkdown v1 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Practical Machine Learning - Peer Graded Assignment

### Summary
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

The goal of you project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

```{r libraries,echo=FALSE}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
```

```{r download,echo=FALSE}
# Check if the dataset exists. If it doesn't, then download.
if (!file.exists("data")) {
  dir.create("data")
}

fileUrlTraining <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv" ##trainingdata
fileUrlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" ##testdata
download.file(fileUrlTraining,destfile="./data/pml-training.csv")
download.file(fileUrlTest,destfile="./data/pml-testing.csv")

dataTraining <- read.csv("data/pml-training.csv",sep=",",header=TRUE,
                         na.strings=c("NA","#DIV/0!",""))

dataTesting <- read.csv("data/pml-testing.csv",sep=",",header=TRUE,
                        na.strings=c("NA","#DIV/0!",""))

```
 
```{r dataAnalysis,echo=FALSE}
dim(dataTraining)
dim(dataTesting)

str(dataTraining$classe)
summary(dataTraining$classe)

```
```{r preprocessing}
inTrain <- createDataPartition(y=dataTraining$classe,p=0.6,list=FALSE)
myTrain <- dataTraining[inTrain, ]
myTest <- dataTraining[-inTrain, ]

dim(myTrain)

## Remove variables that are most NAs.
myTrainClean <- myTrain
for (i in 1:length(myTrain)) {
  if (sum(is.na(myTrain[ , i])) / nrow(myTrain) >= .75) {
    for (j in 1:length(myTrainClean)) {
      if (length(grep(names(myTrain[i]), names(myTrainClean)[j]))==1) {
        myTrainClean <- myTrainClean[ , -j]
      }
    }
  }
}

dim(myTrainClean)

## Remove unneccessary columns
myTrainingNew <- myTrainClean[,8:length(myTrainClean)]

## Remove near zero variables
nearZero <- nearZeroVar(myTrainingNew,saveMetrics=TRUE)
nearZero
```
### Random Decision Forest
```{r randomForest}

## Random Forest
set.seed(123)
modelFit <- randomForest(classe~.,data=myTrainingNew)
print(modelFit)
```
```{r crossValidation}
## Cross Validation
testPrediction <- predict(modelFit,dataTesting,type="class")

trainingResults <- predict(modelFit,myTrainingNew)
trainingAccuracy <- sum(trainingResults==myTrainingNew$classe)/length(trainingResults)
paste("Accuracy on training set =",trainingAccuracy)

validationResults <- predict(modelFit, newdata=myTest)
validationAccuracy <- sum(validationResults==myTest$classe)/length(validationResults)
paste("Accuracy on validation set =",validationAccuracy)

testResults <- predict(modelFit,newdata=dataTesting)
print("Classifications on the test set:");testResults
```
```{r dataVisualization}
heat.tree <- function(tree, low.is.green = FALSE, ...) 
  { 
    # dots args passed to prp
    y <- tree$frame$yval
    if(low.is.green)
      y <- -y
    max <- max(y)
    min <- min(y)
    cols <- rainbow(99, end = .36)[
            ifelse(y > y[1], (y-y[1]) * (99-50) / (max-y[1]) + 50,
                  (y-min) * (50-1) / (y[1]-min) + 1)]
    prp(tree, branch.col = cols, box.col = cols, ...)
  }

treeModel <- rpart(classe ~ ., data=myTrainingNew,method="anova")
heat.tree(treeModel,type=2,varlen=2,fallen.leaves=TRUE)
```